A tool set for random number generation on GPUs in R

### 2022-01-12 pkg/gpuRandom/inst/documents/paper1_2021_5/paper1.Rnw
pkg cite & small fix
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/diseasemapping@3112 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

Simulating Gaussian random fields (GRF's) is computationally intensive due the high dimensionality of the problem, a 100 by 100 grid has 10,000 cells and a 10,000 by 10,000 variance matrix.  Writing $x_1,\dots, x_n$ as point locations (i.e.\ centres of grid cells), the joint distribution of $U=(U_{x_1},\dots,U_{x_n})^\top$ is multivariate Normal (MVN).  An isotropic random field has a variance matrix  with entries which depend on the distances between point locations, and a GRF with geometric anisotoropy has entries depending on the direction as well as the distance.

A geometrically anisotropic GRF with a \citet{matern1960spatial} correlation function is defined as
\begin{aligned*} 
U = &[U(x_1), \ldots, U(x_n)]^\top \sim \text{MVN}(0, \Sigma), \\
\Sigma_{ij} = & \sigma^2 \frac{2^{\kappa-1}}{\Gamma(\kappa)} \left(\sqrt{8\kappa} ||d_{ij}||/\phi \right)^\kappa  K_\kappa\left(\sqrt{8\kappa} ||d_{ij}||/\phi \right),\\
d_{ij} = &
        \left[
          \begin{array}{rr}
            1 & 0 \\
            0 & \omega
          \end{array}
        \right]
        \left[
          \begin{array}{rr}
            \cos(\theta) & -\sin(\theta) \\
            \sin(\theta) & \cos(\theta)
          \end{array}
        \right]
(x_i - x_j)
\end{gather*}
where $K_\kappa(\cdot)$  is a modified Bessel function of the second kind with order $\kappa$ and $\Gamma(\cdot)$ is the standard gamma function.  The model paramaters are:
\begin{itemize}
\item $\sigma$, the (marginal) variance  $\var(U_i) = \sigma$;
\item $\kappa$, the shape parameter controlling the differentiability of $U(\cdot)$;
\item $\phi$, the range parameter controlling how quickly correlation decays with distance in the direction where correlation is strongest;
\item $\omega \geq 1$, a parameter controlling how much faster correlation decays in the direction where correlation is weakest; and 
\item $\theta$, the angle of rotation  required to put the direction of maximum correlation on the x-axis.
\end{itemize}
The standard isotropic Mat\'ern model is obtained by setting $\omega = 1$.
There are several alternative parameterisations of Mat\'ern covariance functions in literature  \citep[see][]{haskard2007anisotropic}. In \pkg{gpuBatchMatrix}, we use the above form with the $\sqrt{8\kappa}$ term as it allows $\phi$ to be interpretable as the distance beyond which correlation is less than 0.14.

The conceptually simplest way of simulating a GRF is to multiply a vector of independent standard Normals by the Cholesky decomposition of $\Sigma$.  This is the direct decomposition method.  
\cite{LiuandLi2019} gives a comprehensive review on seven popular methods for Gaussian random field generation, %which are the turning bands method \citep{matheron_1973}, spectral method \citep{Meja1974OnTS,1972JSV}, matrix decomposition method \citep{davis1987}, Karhunen-Lo\'eve expansion \citep{LoeveM:1978}, moving average method \citep{journel1974, oliver1995moving}, sequential simulation \citep{johnson1987multivariate, gomez1993joint, pebesma2004multivariable}, and local average subdivision \citep{fenton1990simulation}. 
all of these methods other than the direct matrix decomposition method, involve approximations to the random field and have specific requirements on the type of grid or covariance functions. The direct matrix decomposition method is exact, it works for all covariance functions and can generate random fields on arbitrary point locations, and is straightforward to implement.

@@

Other \proglang{R} packages that offer simulation of Gaussian random fields such as \pkg{geoR} \citep{geoR2001}, does not work for large number of locations. The \pkg{RandomFields} \citep{RandomFields2015} package has several different methods for simulation of Gaussian fields, among which the circulant embedding, which is an improved method on direct matrix decomposition, and some variants of the method like the cut-off embedding \citep{gneiting2006fast} are also exact and fast for isotropic Gaussian random fields, however, they work only on rectangular grids.

@@

The following shows an example, in which we simulate on GPU eight Gaussian random fields of Mat\'ern covariance at one time with four sets of parameters, by taking advantage of the GPU-capabilities offered by \pkg{clrng} and \pkg{gpuBatchMatrix} together.  Motivated by the classic Swiss rain dataset, we simluate random fields on a grid of points covering Switzerland.

@@

Figure \ref{fig:maternplot} uses some facilities from the \pkg{mapmisc} package.

>>>>

Simulating Gaussian random fields (GRF's) is computationally intensive due the high dimensionality of the problem, a 100 by 100 grid has 10,000 cells and a 10,000 by 10,000 variance matrix.  Writing $x_1,\dots, x_n$ as point locations (i.e.\ centres of grid cells), the joint distribution of $U=(U_{x_1},\dots,U_{x_n})^\top$ is multivariate Normal (MVN).  An isotropic random field has a variance matrix  with entries which depend on the distances between point locations, and a GRF with geometric anisotoropy has entries depending on the direction as well as the distance.  
A geometrically anisotropic GRF with a \citet{matern1960spatial} correlation function is defined as
\begin{align*}
U &= [U(x_1), \ldots, U(x_n)]^\top \sim \text{MVN}(0, \Sigma), \\
\Sigma_{ij} &= \cov[ U(x_i),U(x_j) ]= \sigma^2 \frac{2^{\kappa-1}}{\Gamma(\kappa)} \left(\sqrt{8\kappa} ||d_{ij}||/\phi \right)^\kappa  K_\kappa\left(\sqrt{8\kappa} ||d_{ij}||/\phi \right),\\
d_{ij} &= 
        \left[
          \begin{array}{rr}
            1 & 0 \\
            0 & \omega
          \end{array}
        \right]
        \left[
          \begin{array}{rr}
            \cos(\theta) & -\sin(\theta) \\
            \sin(\theta) & \cos(\theta)
          \end{array}
        \right]
(x_i - x_j)
\end{align*}
where $K_\kappa(\cdot)$  is a modified Bessel function of the second kind with order $\kappa$ and $\Gamma(\cdot)$ is the standard gamma function.  The model parameters are:
\begin{itemize}
\setlength{\parskip}{0pt}
\item $\sigma$, the (marginal) variance  $\var(U_i) = \sigma$;
\item $\kappa$, the shape parameter controlling the differentiability of $U(\cdot)$;
\item $\phi$, the range parameter controlling how quickly correlation decays with distance in the direction where correlation is strongest;
\item $\omega \geq 1$, a parameter controlling how much faster correlation decays in the direction where correlation is weakest; and 
\item $\theta$, the angle of rotation  required to put the direction of maximum correlation on the x-axis.
\end{itemize}
The standard isotropic Mat\'ern model is obtained by setting $\omega = 1$.
There are several alternative parameterisations of Mat\'ern covariance functions in literature  \citep[see][]{haskard2007anisotropic}. In \pkg{gpuBatchMatrix}, we use the above form with the $\sqrt{8\kappa}$ term as it allows $\phi$ to be interpretable as the distance beyond which correlation is less than 0.14.

The conceptually simplest way of simulating a GRF is to multiply a vector of independent standard Normals by the Cholesky decomposition of $\Sigma$.  This is the direct matrix decomposition method.  
\cite{LiuandLi2019} gives a comprehensive review on seven popular methods for GRF generation, %which are the turning bands method \citep{matheron_1973}, spectral method \citep{Meja1974OnTS,1972JSV}, matrix decomposition method \citep{davis1987}, Karhunen-Lo\'eve expansion \citep{LoeveM:1978}, moving average method \citep{journel1974, oliver1995moving}, sequential simulation \citep{johnson1987multivariate, gomez1993joint, pebesma2004multivariable}, and local average subdivision \citep{fenton1990simulation}. 
all of these methods other than the matrix decomposition method, involve approximations to the random field and have specific requirements on the type of grid or covariance functions. The direct matrix decomposition method is exact, it works for all covariance functions and can generate random fields on arbitrary point locations, and is straightforward to implement.

@@

Other \proglang{R} packages that offer simulation of GRF's such as \pkg{geoR} \citep{geoR2001}, does not work for large number of locations. The \pkg{RandomFields} \citep{RandomFields2015} package has several different methods for simulation of Gaussian fields, among which the circulant embedding, which is an improved method on direct matrix decomposition, and some variants of the method like the cut-off embedding \citep{gneiting2006fast} are also exact and fast for isotropic GRF's, however, they work only on rectangular grids.

@@

The following shows an example, in which we simulate on GPU eight GRF's of Mat\'ern covariance at one time with four sets of parameters, by taking advantage of the GPU-capabilities offered by \pkg{clrng} and \pkg{gpuBatchMatrix} together.  Motivated by the classic Swiss rain dataset, we simluate random fields on a grid of points covering Switzerland.

@@

Figure \ref{fig:maternplot} uses some facilities from the \pkg{mapmisc} \citep{mapmiscPackage} package.
